\acknowledgments{
    I am incredibly grateful to Sumit Gulwani, who first introduced me to the world of research, mentored and advised
    me, believed in my ideas, supported me, and filled our collaboration with indescribable energy, support, and drive.
    Sumit taught me to value simple problem solutions that leave an impact on millions of people, to persevere in the
    face of failure or rejection, and to recognize good research skills.
    Whether in work or in personal life, I knew I could always rely on Sumit's kind words and advice, and for that I
    remain eternally grateful and indebted.

    I have also been blessed to be advised by Zoran Popovi\'c.
    His vision and multi-disciplinary ambitions have always pushed me beyond my comfort zone, helped to keep a bigger
    picture in mind, and motivated to explore further.
    Working with Zoran has been a pleasure, and I will always remember his support, patience, and advice.

    I would like to thank all my co-authors, collaborators, colleagues, and friends who have helped me with these and
    other projects over the course of my graduate education:
    Erik Andersen,
    Rastislav Bodik,
    Marc Brockschmidt,
    Eric Butler,
    Sarah Chasins,
    Loris D'Antoni,
    Kevin Ellis,
    Michael Ernst,
    Elena Glassman,
    Maxim Grechkin,
    Bj\"orn Hartmann,
    Pushmeet Kohli,
    Ranvijay Kumar,
    Vu Le,
    Mark Marron,
    Mika\"el Mayer,
    Saswat Padhi,
    Daniel Perelman,
    Mark Plesko,
    Mohammad Raza,
    Eleanor O'Rourke,
    Danny Simmons,
    Rishabh Singh,
    Adam M. Smith,
    Gustavo A. Soares,
    Emina Torlak,
    Abhishek Udupa,
    Luke Zettlemoyer,
    and
    Ben Zorn.

    Last but not least, I want to thank my parents and Sasha for their unending love and support, for words of pride and
    encouragement, and for tolerating my stressful schedule and mood swings.
    Without you, none of this would ever happen.
}
