\section{Witness Functions}
\label{sec:prose:wf}
As described above, a \emph{witness function} is a generalization of inverse semantics for an operator.
In other words, it is a \emph{problem reduction logic}, which deduces a necessary (or sufficient) spec on the operator's
parameters given a desired spec on the operator's output.
In this section, I define witness functions formally and give various examples of their usages in the existing PBE
technologies.

\begin{defn}[Witness function]
    Let $F\left(N_1,\dots,N_k\right)$ be an operator in a DSL $\dsl$.
    A \emph{witness function} of $F$ for $N_j$ is a function $\omega_{j}\left(\spec\right)$ that deduces a
    \emph{necessary} spec $\spec_j$ on $N_j$ given a spec $\spec$ on $F\left(N_1, \dots, N_k\right)$.
    Formally, $\omega_j\left(\spec\right) = \spec_j$ iff the following implication holds:\footnote{All free variables
        are universally quantified unless otherwise specified.}
    \begin{equation*}
        F\left(N_1, \dots, N_k\right) \models \spec \qquad\Longrightarrow\qquad N_j \models \spec_j.
    \end{equation*}
\end{defn}

\begin{defn}[Precise witness function]
    A witness function $\omega_j$ of $F\left(N_1, \dots, N_k\right)$ for $N_j$ is \emph{precise} if its deduced spec is
    \emph{necessary and sufficient}.
    Formally, $\omega_j\left(\spec\right) = \spec_j$ is precise iff
    \begin{equation*}
        N_j \models \spec_j \qquad\Longleftrightarrow\qquad
        \exists\, N_1, \dots, N_{j-1}, N_{j+1}, \dots, N_k\colon\ F\left(N_1, \dots, N_k\right) \models \spec.
    \end{equation*}
\end{defn}

\begin{defn}[Conditional witness function]
    A \emph{(precise) conditional witness function} of $F\left(N_1, \dots, N_k\right)$ for $N_j$ is a function
    $\omega_j\left(\spec \assuming N_{k_1} = v_1, \dots, N_{k_s} = v_s\right)$ that deduces a necessary (and sufficient)
    spec $\spec_j$ on $N_j$ given a spec $\spec$ on $F\left(N_1, \dots, N_k\right)$ under the assumption that a subset
    of other parameters $N_{k_1}, \dots, N_{k_s}$ of $F$ (called \emph{prerequisites}) have fixed values $v_1, \dots,
    v_k$.
    Formally, $\omega_j\left(\spec \assuming N_t = v_t\right) = \spec_j$ iff the following implication holds:
    \begin{equation*}
        F\left(N_1, \dots, N_{t-1}, v_t, N_{t+1}, \dots, N_k\right) \models \spec
        \qquad\Longrightarrow\qquad
        N_j \models \spec_j.
    \end{equation*}
\end{defn}

\input{wf-list}

\begin{example}
    \Cref{tbl:wfs:flashfill} shows the witness functions for all $\ffdsl$ operators from \Cref{fig:dsl:flashfill}:
    \begin{itemize}[nosep]
        \item A \dslinline|Concat($atom$, $transform$)| expression returns $w$ iff $atom$ returns some prefix of $w$.
            In addition, assuming that $atom$ returns $v$, $transform$ mush return the remaining suffix of~$w$ after the
            end of $v$.
        \item A \dslinline|ConstStr($s$)| expression returns $w$ iff $s$ is equal to $w$.
        \item An expression ``\dslinline|let $x$ = std.Kth($inputs$, $k$) in $\dots$|'' returns $w$ iff $x$ is bound to
            an element of $inputs$ that has $w$ as a substring.
        \item A \dslinline|Substring($x$, $pp$)| expression returns $w$ (assuming that $x$ returns $v$) iff $pp$ returns
            a position span of any occurrence of $w$ in $v$ as a substring.
        \item An \dslinline|AbsPos($x$, $k$)| expression returns $c$ (assuming that $x$ returns $v$) iff $k$ is equal to
            either~$c$ or $c-|v|-1$ (since $k$ may represent a left or right offset depending on its sign).
        \item An expression \dslinline|RegexPos($x$, $rr$, $k$)| returns $c$ (assuming that $x$ returns $v$) iff $rr$ is
            equal to any pair of regular expressions that matches the boundaries of position $c$ in the string~$v$.
            In addition, assuming that $rr$ is equal to $\langle r_1, r_2\rangle$, $P$ returns $c$ iff $k$ is equal to
            a index of $c$ (from the left or right) among all matches of $\langle r_1, r_2\rangle$ in $v$.
    \end{itemize}
    \label{ex:wf:flashfill}
\end{example}

\begin{example}
    \Cref{tbl:wfs:prose} shows the witness functions for various operators from the standard library of PROSE that are
    used in $\ffdsl$ and $\fedsl$:
    \begin{itemize}[nosep]
        \item A \dslinline|Kth($xs$, $k$)| expression returns $w$ (given that $xs$ returns $\values$) iff $k$ is an
            index of some occurrence of $w$ in $\values$.
        \item If \dslinline|Pair($p_1$, $p_2$)| returns $\langle v_1, v_2\rangle$, then $p_j$ returns $v_j$.
            Note that this witness function is imprecise, since it restricts only a single parameter ($p_1$ or $p_2$).
        \item A \dslinline|Map($F$, $L$)| expression returns a list that starts with a sublist $\vec{\ell}$ as a prefix
            (given that $L$ returns $\vec{v}$) iff $F$ is any $\lambda$-function that maps the first $|\vec{\ell}|$
            elements of $\vec{v}$ to the corresponding elements of $\vec{\ell}$.
            The fact that $L$ returns $\vec{v}$ is usually established thanks to a corresponding domain-specific witness
            function for $L$ (defined by a DSL designer for a particular instantiation of \texttt{Map}).
        \item A \dslinline|Filter($P$, $L$)| expression returns a list that contains $\vec{\ell}$ as a sublist iff the
            result of $L$ also contains $\vec{\ell}$ as a sublist.
            In addition, assuming that $L$ returns a list $\vec{v}$, $P$ must be any $\lambda$-function that maps all
            the elements of $\vec{v}$ that are also present in $\vec{\ell}$ to \texttt{true}, and the rest of the
            elements to \texttt{false}.
        \item A \dslinline|FilterInt($i$, $k$, $L$)| expression returns a list that contains $\vec{\ell}$ as a sublist
            iff the result of $L$ also contains $\vec{\ell}$ as a sublist.
            In addition, assuming that $L$ returns a list $\vec{v}$, the initial value expression $i$ must evaluate to
            the index of $\ell_1$ in $\vec{v}$, and the step expression $k$ must evaluate to some divisor of the GCD of
            the gaps between the consecutive occurrences of the elements of $\vec{\ell}$ in $\vec{v}$.
    \end{itemize}
\end{example}

\begin{example}
    As mentioned above, a witness function for the parameter $L$ of \dslinline|Map($F$, $L$)| must be defined
    specifically for every instantiation of \texttt{Map} in a given DSL.
    For instance, the FlashExtract DSL $\fedsl$ contains three \texttt{Map} instantiations: \texttt{LinesMap},
    \texttt{StartSeqMap}, and \texttt{EndSeqMap} (as defined in \Cref{fig:dsl:flashfill}).
    Their respective witness functions $\omega_L$ for the same \emph{prefix spec} $\state \tospec \mathbf{?} \sqsupseteq
    \vec{v}$ are shown below:

    \vspace{0.5\baselineskip}
    \begin{tabu}{l>{\hspace{0.1cm}}X[$l]}
        \toprule
        \textbf{Operator} & \textbf{Output spec} \\
        \midrule
        \texttt{LinesMap} & \begin{aligned}
            &\state \tospec \mathbf{?} \sqsupseteq \bigl[ \ell_1, \dots, \ell_{|\vec{v}|} \bigr] \quad\text{ where } \\
            &\quad \ell_i \text{ is a line of the input document } \state[doc] \text{ that contains the region } v_i
        \end{aligned} \\
        \texttt{StartSeqMap} &
            \state \tospec \mathbf{?} \sqsupseteq \bigl[p_1 \mid \langle p_1, p_2 \rangle \in \vec{v}\bigr] \\
        \texttt{EndSeqMap} &
            \state \tospec \mathbf{?} \sqsupseteq \bigl[p_2 \mid \langle p_1, p_2 \rangle \in \vec{v}\bigr] \\
        \bottomrule
    \end{tabu}
\end{example}

\paragraph{}
Most witness functions are domain-specific w.r.t.  the operator that they characterize.
However, once formulated in a module for a domain such as substring extraction, they can be reused by any DSL.
In our example, witness functions for most operators in \Cref{tbl:wfs:prose} (namely, all except \texttt{Map}) do not
depend on the domain of their parameters, and are therefore formulated generically, for any DSL.
Witness functions in \Cref{tbl:wfs:flashfill} hold only for their respective operators, but they do
not depend on the rest of the DSL in which these operators are used, provided the operator semantics is conformant with
its (strongly-typed) signature.
This property allows us to define witness functions as generally as possible in order to reuse the corresponding
operators in any conformant DSL.
