\chapter{Interactive Program Synthesis}
\label{ch:interactive}

The key challenge of PBE and, more generally, programming with under-specifications, is \emph{intent ambiguity}.
Examples are an inherently ambiguous form of specification.
A typical industrial DSL usually may contain up to $10^{20}$ programs that are consistent with a given input-output
example~\cite{singh2012synthesizing}.
The underlying PBE system might end up synthesizing an unintended program that is consistent with the examples provided
by the user but does not generate the intended results on some other inputs that the user cares about.
In \citeyear{pbd-fail}, \citeauthor{pbd-fail} presented a critical discussion of PBE systems noting that adoption
of PBE systems is not yet widespread, and proposing that this is mainly due to lack of usability
and confidence in such systems~\cite{pbd-fail}.
We have observed the issues of confidence, transparency, and debuggability to arise in most interactions with PBE-based
synthesis systems.

In the past, intent ambiguity in PBE has been primarily handled by imposing a sophisticated ranking on the
DSL~\cite{cav:ranking}.
While ranking goes a long way in avoiding undesirable interpretations of the user's intent, it is not a complete
solution.
For example, FlashFill is designed to cater to users that care not about the program but about its behavior on the small
number of input rows in the spreadsheet.
Such users can simply eye-ball the outputs of the synthesized program and provide another example if they are incorrect.
However, this becomes much more cumbersome (or impossible) with a larger spreadsheet.%
\footnote{John Walkenbach, famous for his Excel textbooks, labeled \ff as a ``controversial'' feature.
    He wrote: \textit{``It's a great concept, but it can also lead to lots of bad data. I think many users will look at
        a few ``flash filled'' cells, and just assume that it worked. But my preliminary tests leads me to this
        conclusion: Be very careful.''}~\cite{walkenbach:controversial}}
Moreover, inspecting the synthesized program directly also does not establish enough confidence in it even if the user
knows programming.
Two main reasons for this are (i) program readability,%
\footnote{Stephen Owen, a certified MVP (``Most Valued Professional'')
  in Microsoft technologies, said the following of a program
  synthesized by \fe: \textit{``If you can understand this, you’re a better person than I am.''}~\cite{mvp:complaint}}
and (ii) the users' uncertainty in the desired intent due to hypothetical unseen corner cases in the data.

Due to ambiguity of intent in PBE, the standard user interaction model in this setting is for the user to provide
constraints \emph{iteratively} until the user is satisfied with the synthesized program or its behavior on the known
inputs.
However, most work in this area, including FlashFill and FlashExtract, has not been formally modeled as an iterative
process.
In this work, I describe our interactive formulation of program synthesis that leverages the inherent iterative nature
of synthesis from under-specifications, as well as multiple novel techniques for intent disambiguation and user
interaction models that leverage the innate strengths of this interactive formulation.

In \Cref{sec:interactive:problem}, I formally introduce the problem of \emph{interactive program synthesis}, wherein an
application-specific \emph{synthesis session state} is kept after each iteration.
This state is used to improve the transparency, performance, and debuggability of the synthesis process via various
means, including proactive feedback to the user, directing the user toward better examples, and speeding up the next
synthesis iterations incrementally.

In \Cref{sec:interactive:models}, I describe our exploration and evaluation of user interaction models that aim to
improve the transparency of the iterative synthesis process.
Our proposed two novel interaction models alleviate above\hyp{}mentioned transparency concerns
by exposing more information to the user in a form that can be easily understood and acted upon.
These models help resolve ambiguity in the example\hyp{}based specification, thereby increasing
user's trust in the results produced by the PBE engine.
The first model, \emph{Program Navigation}, allows the user to navigate
between all programs synthesized by the underlying PBE engine (as opposed to displaying only the
top-ranked program) and to pick one that is intended.
We leverage the implicit sharing present in a VSA~(\Cref{ch:vsa}) to create a navigational interface that allows the
user to select from different ranked choices for various parts of the top-ranked program.
Furthermore, these programs are paraphrased in English for easy readability.
The second model, \emph{Conversational Clarification}, is based on active learning.
In it, the system asks questions to the user to resolve ambiguities in the user's
specification with respect to the available test data.
These questions are generated after the PBE engine has synthesized multiple programs that are consistent with the
user-provided examples.
The system executes these multiple programs on the test data to identify any discrepancies in the execution and uses
that as the basis for asking questions to the user.
The user responses are used to refine the initial example-based specification and the
process of program synthesis is repeated.

\Cref{sec:interactive:feedback} builds on the ideas of Conversational Clarification, generalizing it into a universal
formal technique that is applicable to an arbitrary interactive synthesis process -- \emph{feedback-based synthesis}.
It builds on our interactive problem definition from \Cref{sec:interactive:problem}.
I present a method for leading the user toward the examples with maximum disambiguation potential, and discuss our
usage of feedback-based synthesis in PROSE-based applications.

Finally, in \Cref{sec:interactive:incremental}, I show how the same problem formulation can be used to improve the
performance of the synthesis process.
The standard PBE model requires the user to refine her intent in iterative rounds by providing additional constraints on
the current candidate program.
The standard approach has been to re-run the synthesizer afresh with the conjunction of the original constraints and the
new constraints.
In this work, I introduce an alternative technique: leveraging the \emph{interpretation of previously learned VSA as a
sub-language~(\Cref{sec:vsa:language})} to carry out the new round of synthesis on a smaller program space.

\input{problem}
\input{models}
\input{feedback}
\input{incremental}
